Generalized measurement invariance tests with application to factor analysis

### 2012-02-11 www/paper.Rnw
Minor wording changes to the revised paper, based on final read-through.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@103 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\abstract{
  The issue of measurement invariance commonly arises in factor-analytic
  contexts, with methods for assessment including likelihood ratio tests,
  Lagrange multiplier tests, and Wald tests. These tests all require advance
  definition of the number of groups, group membership, and offending model
  parameters. In this paper, we construct tests of measurement invariance based on
  stochastic processes of casewise derivatives of the likelihood function. These
  tests can be viewed as generalizations of the Lagrange multiplier test, and they
  are especially useful for: (1)~isolating specific parameters affected by
  measurement invariance violations, and (2)~identifying subgroups of individuals
  that violated measurement invariance based on a continuous auxiliary variable. 
  The tests are presented and illustrated in detail, along with an application
  to a study of stereotype threat and simulations examining the tests'
  abilities in controlled conditions.
}

@@

Many expositions of factor analysis utilize the likelihood for the sample
covariance matrix (instead of the likelihood of the individual observations $\bm{x}_i$),
employing a Wishart distribution. However, in that approach the sample covariance
matrix can also be disaggregated to the sum of its individual-specific contributions, 
leading essentially to a likelihood like \eqref{eq:mvndensity} with multivariate normal
observations $\bm{x}_i$. This situation is also generally encountered in structural
equation models with missing data (e.g., \citeNP{Wot00}).

@@

For assessing measurement invariance in psychometric models, the major limitation of
the three tests is that the potential subgroups have to be known in advance. 
Even if the variable $V$ w.r.t.\ which the violation of invariance occurs is
known, the threshold $\nu$ from (\ref{eq:h1*}) is often unknown in practice.
For example, if $V$ represents yearly income, there are many possible
values of $\nu$ that could be used to divide individuals into poorer and richer
groups.  The ultimate $\nu$ that we
choose could potentially impact our
conclusions about whether or not a scale is measurement invariant, 
in the same way that dichotomization of continuous variables impacts
general psychometric analyses \cite{MacZha02}.

@@

If a measurement invariance violation is detected by any of the tests,
one may want to incorporate it into the model to account for it. The
procedure for doing this  typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
discussed above often prove helpful in determining a suitable parameterization.
In particular, one approach that is often employed in practice
involves adoptions of a model
with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
single threshold case). In the multiple threshold case, their location can
be determined by maximizing the corresponding segmented log-likelihood over all
possible combinations of thresholds (\citeNP{ZeiSha10}, adapt a dynamic programming
algorithm to this task). For the single threshold case, this reduces to
maximizing the segmented log-likelihood
\begin{equation} \label{eq:seglik}
  \ell(\hat {\bm \theta}^{({A)}}; x_1, \dots, x_m) ~+~ \ell(\hat {\bm \theta}^{({B)}}; x_{m+1}, \dots, x_n)
\end{equation}
over all values of $m$ corresponding to possible thresholds $\nu$ (such that $v_m \le \nu$ and
$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR~statistic
from (\ref{eq:lr}) (with some minimal subgroup size typically imposed).

@@

As mentioned previously, the double-max test is advantageous because
it yields information about individual parameters violating measurement invariance.  
That is, it allows us to examine whether or not individual parameters' 
cumulative score processes lead one to reject the hypothesis of
measurement invariance.
Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
three verbal factor loadings and the top left panel of Figure~\ref{fig:cusumex} shows
the same processes aggregated over the three parameters. In both graphics, horizontal lines reflecting the
same critical value at $\alpha= 0.05$. The figure
shows that the third parameter (i.e., $\lambda_{31}$) crosses the
dashed line, so we would conclude a measurement 
invariance violation with respect to age for the third
verbal test.  The fact that the cumulative score process for the first
and second loadings did not achieve the critical value represents a type~II
error, which implicitly brings into question the tests' power.
This is an example of a situation where the double-max test ``wastes''
power, as it cannot make use of the fact that multiple model
parameters change simultaneously.  We
generally address the issue of power in the simulations below.

@@

\subsection{Method}
WDH tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step~5b (see pp.~703--704 of their paper), which involved
across-group restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.    The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
(comparative fit index) statistics.  While the model included four
groups as described above, we focus only on the submodel for the minority
group that received stereotype primes, which is pictured in Figure~\ref{fig:wpath}.  In the figure, paths with dashed lines and red parameters signify group-specific parameters.  These include the factor mean $\eta$, numerical factor loading $\lambda_{\text{num}}$, numerical intercept $\mu_{\text{num}}$, and numerical unique variance $\psi_{\text{num}}$.  Other parameters (not displayed) were restricted to be equal across all four groups, with the exception of the factor variance.  This parameter was restricted to be equal within both minority groups, and separately within both majority groups.

@@

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change. That is, we
have the ability to  test various subsets of  parameters.  For the WDH data, we
first test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i, \text{num}}\ \eta_{i}\ \mu_{i, \text{num}}\ \psi_{i, \text{num}}) = 
      (\lambda_{0, \text{num}}\ \eta_{0}\ \mu_{0, \text{num}}\ \psi_{0, \text{num}}),\
      i = 1,\ldots, n_{\text{MSP}},
\end{equation}
where
$\lambda_{i, \text{num}}$ is the factor loading on the numerical scale,
$\eta_{i}$ is student $i$'s factor mean,
$\mu_{i, \text{num}}$ is student $i$'s intercept on the numerical scale, and
$\psi_{i, \text{num}}$ is student $i$'s unique variance on the numerical scale.
The students are ordered according to GPA (i.e.,  in this application the variable $V$),
and the  hypothesis is taken only with respect to the $n_{\text{MSP}}$ individuals in the
``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states  that these four group-specific
model parameters are invariant with respect to GPA.

@@

The double-max test for the hypothesis~\eqref{eq:exh0} is shown in
Figure~\ref{fig:cusumgpa}, displaying the cumulative score processes
for each of the four group-specific parameters across GPA along
with horizontal lines representing the critical value for $\alpha= 0.05$.
This shows that the ``minority, stereotype prime'' group lacks invariance
with respect to GPA,
because one of the processes crosses its boundaries. Furthermore, as this
process pertains to the variance $\psi_\text{num}$ of the numerical scale
(bottom panel) it can be concluded that the invariance violation is associated
with this parameter. However, also the factor loading for the numerical scale
$\lambda_\text{num}$ and the factor mean $\eta$ exhibit increased fluctuation
albeit they are just non-significant at level $\alpha = 0.05$. Only the process
associated with the intercept $\mu_\text{num}$ just shows moderate (and hence
clearly non-significant) fluctuation across GPA.

@@

\subsection{Categorical Auxiliary Variable}
One issue that was largely unaddressed in this paper involved the use
of categorical $V$ to study measurement invariance.  In this case,
groups are 
already specified in advance, and so traditional methods for fixed
subgroups (i.e., LR, Wald, and LM tests) may suffice.
Furthermore, we can also obtain an LM-type statistic from the
framework developed here. Assume the observations are divided into
$C$ categories $I_1, I_2, \ldots, I_C$. Then, the increment of
the cumulative score process $\Delta_{I_c} {\bm B}(\hat {\bm \theta})$
within each category is just the sum of the corresponding scores.
In somewhat sloppy notation:
\begin{equation}
    \label{eq:catsum}
    \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i)
%% \\
%%    X^2 & = & \sum_{c = 1}^C n_c^{-1} n^{-1} || \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ||^2
\end{equation}
This results in a $C \times k$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $C \times k$ matrix and employing a weighted squared sum of
the entries in the column to obtain a $\chi^2$-distributed statistic with
$(C-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.  In addition to categorical $V$, this framework may 
be useful for ordinal $V$ (or continuous $V$ with many ties).
In this situation, one could also adapt the statistics
\eqref{eq:dmax}, \eqref{eq:cvm}, and \eqref{eq:maxlm} computed from the same cumulative score
process as usual. The only required modification is that the statistics
should not depend on the process's values ``within'' a category (or group
of ties). This is easily achieved by taking the maximum (or sum) not over
all observations $i = 1, \dots, n$, but over only those $i$ at the ``end'' of a
category.
%% If $i_j$ is the cumulative frequency of category $j$ or lower
%% (with $j = 1, \dots, C$), then one could take the outer maximum in
%% \eqref{} over $i = i_1, i_2, \dots, i_{C-1}$.

>>>>

\abstract{
  The issue of measurement invariance commonly arises in factor-analytic
  contexts, with methods for assessment including likelihood ratio tests,
  Lagrange multiplier tests, and Wald tests. These tests all require advance
  definition of the number of groups, group membership, and offending model
  parameters. In this paper, we construct tests of measurement invariance based on
  stochastic processes of casewise derivatives of the likelihood function. These
  tests can be viewed as generalizations of the Lagrange multiplier test, and they
  are especially useful for: (1)~identifying specific parameters impacted by
  measurement invariance violations, and (2)~identifying subgroups of individuals
  that violated measurement invariance based on a continuous auxiliary variable. 
  The tests are presented and illustrated in detail, including an application
  to a study of stereotype threat and simulations examining the tests'
  abilities in controlled conditions.
}

@@

Many expositions of factor analysis utilize the likelihood for the sample
covariance matrix (instead of the likelihood of the individual observations $\bm{x}_i$),
employing a Wishart distribution. However, in that approach the sample covariance
matrix can also be disaggregated to the sum of its individual-specific contributions, 
leading essentially to a likelihood like \eqref{eq:mvndensity} with multivariate normal
observations $\bm{x}_i$. This situation is similar to that encountered in structural
equation models with missing data (e.g., \citeNP{Wot00}).

@@

For assessing measurement invariance in psychometric models, the major limitation of
the three tests is that the potential subgroups have to be known in advance. 
Even if the scales are known to violate measurement invariance w.r.t.\ the variable $V$, the threshold $\nu$ from (\ref{eq:h1*}) is often unknown in practice.
For example, if $V$ represents yearly income, there are many possible
values of $\nu$ that could be used to divide individuals into poorer and richer
groups.  The ultimate $\nu$ that we
choose could potentially impact our
conclusions about whether or not a scale is measurement invariant, 
in the same way that dichotomization of continuous variables impacts
general psychometric analyses \cite{MacZha02}.

@@

If a measurement invariance violation is detected by any of the tests,
one may want to incorporate it into the model to account for it. The
procedure for doing this  typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
discussed above often prove helpful in determining a suitable parameterization.
In particular, one approach that is often employed in practice
involves adoption of a model
with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
single threshold case). In the multiple threshold case, their location can
be determined by maximizing the corresponding segmented log-likelihood over all
possible combinations of thresholds (\citeNP{ZeiSha10}, adapt a dynamic programming
algorithm to this task). For the single threshold case, this reduces to
maximizing the segmented log-likelihood
\begin{equation} \label{eq:seglik}
  \ell(\hat {\bm \theta}^{({A)}}; x_1, \dots, x_m) ~+~ \ell(\hat {\bm \theta}^{({B)}}; x_{m+1}, \dots, x_n)
\end{equation}
over all values of $m$ corresponding to possible thresholds $\nu$ (such that $v_m \le \nu$ and
$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR~statistic
from (\ref{eq:lr}) (with some minimal subgroup size typically imposed).

@@

We previously noted that the double-max test is advantageous because
it yields information about individual parameters violating measurement invariance.  
That is, it allows us to examine whether or not individual parameters' 
cumulative score processes lead one to reject the hypothesis of
measurement invariance.
Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
three verbal factor loadings and the top left panel of Figure~\ref{fig:cusumex} shows
the same processes aggregated over the three parameters. In both graphics, horizontal lines reflect the
same critical value at $\alpha= 0.05$. The figure
shows that the third parameter (i.e., $\lambda_{31}$) crosses the
dashed line, so we would conclude a measurement 
invariance violation with respect to age for the third
verbal test.  The fact that the cumulative score process for the first
and second loadings did not achieve the critical value represents a type~II
error, which implicitly brings into question the tests' power.
% This is an example of a situation where the double-max test ``wastes''
% power, as it cannot make use of the fact that multiple model
% parameters change simultaneously.
We
generally address the issue of power in the simulations below.

@@

\subsection{Method}
WDH tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step~5b (see pp.~703--704 of their paper), which involved
across-group restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.    The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
(comparative fit index) statistics.  While the model included four
groups as described above, we focus only on the submodel for the minority
group that received stereotype primes, which is pictured in Figure~\ref{fig:wpath}.  In the figure, paths with dashed lines and red parameters signify group-specific parameters.  These include the factor mean $\eta$, numerical factor loading $\lambda_{\text{num}}$, numerical intercept $\mu_{\text{num}}$, and numerical unique variance $\psi_{\text{num}}$.  Other parameters (not displayed) were restricted to be equal across all four groups, with the exception of the factor variance.  The factor variance was restricted to be equal within both minority groups, and separately within both majority groups.

@@

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change: we
have the ability to  test various subsets of  parameters.  For the WDH data, we
first test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i, \text{num}}\ \eta_{i}\ \mu_{i, \text{num}}\ \psi_{i, \text{num}}) = 
      (\lambda_{0, \text{num}}\ \eta_{0}\ \mu_{0, \text{num}}\ \psi_{0, \text{num}}),\
      i = 1,\ldots, n_{\text{MSP}},
\end{equation}
where
$\lambda_{i, \text{num}}$ is the factor loading on the numerical scale,
$\eta_{i}$ is student $i$'s factor mean,
$\mu_{i, \text{num}}$ is student $i$'s intercept on the numerical scale, and
$\psi_{i, \text{num}}$ is student $i$'s unique variance on the numerical scale.
The students are ordered according to GPA (i.e.,  in this application the variable $V$),
and the  hypothesis is taken only with respect to the $n_{\text{MSP}}$ individuals in the
``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states  that these four group-specific
model parameters are invariant with respect to GPA.

@@

The double-max test for the hypothesis~\eqref{eq:exh0} is shown in
Figure~\ref{fig:cusumgpa}, displaying the cumulative score processes
for each of the four group-specific parameters across GPA along
with horizontal lines representing the critical value for $\alpha= 0.05$.
This shows that the ``minority, stereotype prime'' group lacks invariance
with respect to GPA,
because one of the processes crosses its boundaries. Furthermore, as this
process pertains to the variance $\psi_\text{num}$ of the numerical scale
(bottom panel) it can be concluded that the invariance violation is associated
with this parameter.  Both the factor loading for the numerical scale
$\lambda_\text{num}$ and the factor mean $\eta$ also exhibit increased fluctuation, 
although they are just non-significant at level $\alpha = 0.05$. Only the process
associated with the intercept $\mu_\text{num}$ shows moderate (and hence
clearly non-significant) fluctuation across GPA.

@@

\subsection{Categorical Auxiliary Variable}
One issue that was largely unaddressed in this paper involved the use
of categorical $V$ to study measurement invariance.  In this case,
groups are 
already specified in advance, and so traditional methods for fixed
subgroups (i.e., LR, Wald, and LM tests) may suffice.
However, we can also obtain an LM-type statistic from the
framework developed here. Assume the observations are divided into
$C$ categories $I_1, I_2, \ldots, I_C$. Then, the increment of
the cumulative score process $\Delta_{I_c} {\bm B}(\hat {\bm \theta})$
within each category is just the sum of the corresponding scores.
In somewhat sloppy notation:
\begin{equation}
    \label{eq:catsum}
    \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i).
%% \\
%%    X^2 & = & \sum_{c = 1}^C n_c^{-1} n^{-1} || \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ||^2
\end{equation}
This results in a $C \times k$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $C \times k$ matrix and employing a weighted squared sum of
the entries in the column to obtain a $\chi^2$-distributed statistic with
$(C-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.  In addition to categorical $V$, this framework may 
be useful for ordinal $V$ (or continuous $V$ with many ties).
In this situation, one could also adapt the statistics
\eqref{eq:dmax}, \eqref{eq:cvm}, and \eqref{eq:maxlm} computed from the same cumulative score
process as usual. The only required modification is that the statistics
should not depend on the process's values ``within'' a category (or group
of ties). This is easily achieved by taking the maximum (or sum) not over
all observations $i = 1, \dots, n$, but over only those $i$ at the ``end'' of a
category.
%% If $i_j$ is the cumulative frequency of category $j$ or lower
%% (with $j = 1, \dots, C$), then one could take the outer maximum in
%% \eqref{} over $i = i_1, i_2, \dots, i_{C-1}$.
